### 2011-07-12 Paper/Paper.tex
Updated paper, reformulated relative features to correspond to Widmer

Structure seems to be directly related to expression and it seems logical that a performance rendering system should have some notion of usical structure. The system presented here will use structure exclusively to generate performances and will completely ignore note level expression. The expressiveness of the performances it generates is therefore limited. In section XX we will describe how this system could be integrated with a note level performance rendering system to hopefully produce some variant of YQX that doesn't need explicit rules to produce expressive performances.

@@

Although the YQX does have some notion of structure\footnote{One of the note features is distance to nearest point of closure}, it still only looks at note level expression. The authors admit that the first simple version of the system ``tended to sometimes produce unstable, `nervous' sounding performances''. We see this as a symptom of a note level expression based system.

In this thesis, we propose a structure based performance rendering (SBPR) system. The system presented here ignores note level expression. Instead we will only try to predict constituent level expression. The assumption is that this kind of expression really exists in performances and that is different and independend from note level structure. We think that a constituent level system also corresponds better to how actual human performers play music. A performance redering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.

The system will be similar to YQX in a number of ways, but instead of predicting expression per note, it will predict expression per constituent. Every constituent will be played with consistend expression. A number of score features defined per constituent will be used to predict epressive parameters of the constituent.

The simplification of ignoring note level expression is of course unjustified and severly limits the expressiveness of the performances that can be generated. However, if succesful, the resulting performances will clearly demonstrate the phenomenon of constituent level expression. It should be possible to extend the system to incorporate note level expression, section \ref{integration} will further explore this possibility.

The succes of a structure based performance rendering (SBPR) system depends largely on two factors. The ability to generate musically meaningful parse trees of a piece and the ability to accurately characterize the individual constituents and their relations with other constituents. The following two sections address these issues.

@@

\section{The Delta Framework}

In his Phd thesis \ref{markwin}, Markwin van der Berg introduces a formal way of parsing music into parsetrees: the delta framework. He relates his work to the work of Lerdahl and Jackendoff \ref{lerdahl1996generative} but claims to have found a more general approach.

@@

\[M = n_i, n_{i+1}, \cdots, n_j\]

We can assign features to these notes.

\subsection{Extracting structure}

@@

\section{Constituent Features}

We can now convert a piece of music into a serie of constituents. These constituents will be used to predict expression so we must be able to charaterize them in a way correlates with the way they are performed. Analogous to YQX we are looking for the \textit{context} of the constituent.

@@

\section{Targets}

@@

\section{Performance Rendering}
\begin{itemize}
\item Extract melody
\item Extract scorefeatures from melody
\item lookup notes in deviation, ignore missing notes
\item Extract expressive and non-expressive melody and extract expression features
\item Perform notes other than melody notes the same as the nearest melody note
\end{itemize}
\subsection{Dataset and representation}
\begin{itemize}
\item attack release
\item tempo curves
\end{itemize}
\subsection{Dealing with polyphony}
\subsection{Discretization}

@@

\subsubsection*{The delta framework}

\subsection{Performance Model}

\subsubsection{Expressive timing} Honing and Desain, Honing and timmers, Honing.
\subsection{Learning}

\section{Method}
\section{Evaluation and Results}
The proposed system actually requires more training data to do meaningful statistics. However, since expressive interpretation can be 
\subsection{Subjective Listening}
\subsection{Correlation}
\section{Integration with Note Level Performance Rendering}

@@

Unfortunately we do not have acces to the large dataset that YQX uses. The dataset we use is smaller. Since we do not learn per note but per group of notes, the impact of a smaller dataset is even larger. We have discretize into rather large bins for this reason, resulting in cartoonistisc performances.

@@

\begin{itemize}
\item Top notes is not always the melody, musical attention
\item Bass and harmony shouldn't be played with the same expression as melodynotes
\end{itemize}

\section{Future Work}
\begin{itemize}
\item Use loudness and tempo direction instead of averages (requires more data)
\item Generalize approach to incorporate hierarchical structure, let structure extend to note level, so note level expression and structurelevel expression become integrated
\end{itemize}
\subsection{Repetition}
A notion of repetition and similarity would certainly improve the system. Repetition is a very good indicator of constituent breaks. Repetition and similarity could also be used to improve expressiveness of performances. It is probably telling when a phrase is repeated three times and then slightly altered the fourth time. Although finding similarity and musically significant repetition is a subject of its own the delta framework could help to define repetition arbitrarily of transposition or rhythm. A list of pitch deltas can for example be used to detect repetition independend of transposition and a list of duration deltas can be used to detect repetition of rhythm independend of the notes used.

>>>>

The system presented here will use structure exclusively to generate performances and will completely ignore note level expression. Section \ref{approach} will describe the system as a whole. What we mean by structure level expression is clarified in section \ref{structure}. The approach for extracting structure is described in section \ref{deltaframework}.

The expressiveness of the performances it generates is therefore limited. In section XX we will describe how this system could be integrated with a note level performance rendering system to hopefully produce some variant of YQX that doesn't need explicit rules to produce expressive performances.

@@

Although the YQX does have some notion of structure\footnote{One of the note features is distance to nearest point of closure}, expression is still only predicted per note. The authors admit that the first simple version of the system ``tended to sometimes produce unstable, `nervous' sounding performances''. We see this as a symptom of a note level expression based system.

In this thesis, we propose a structure based performance rendering (SBPR) system. The system presented here ignores note level expression. Instead we will try to predict only constituent level expression. The assumption is that this kind of expression really exists in performances and that is different and independend from note level structure. We think that a constituent level system also corresponds better to how actual human performers play music. A performance redering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.

%\subsection{Structure Based Performance Rendering}

The system will be similar to YQX in a number of ways, but instead of predicting expression per note, it will predict expression per constituent. Every constituent will be played with consistend expression, the articulation, dynamics and tempo change only at constituent breaks.

We use a corpus that contains performances and corresponding scores of Western classical piano music. Every note in every performance has been associated with the corresponding score note so we can see exactly how every note in the score was played. The performances are of high quality and played by famous pianists. See section \ref{corpus} for more details on the corpus.

A structural analysis is used to derrive hierarchical structure of every score in the corpus, however, to keep the system simple we will only use this structural analysis to create a segmentation of the score into constituents. After segmentation, four score features, two of which are direct generalizations of YQX's score features, are extracted for each constituent.

So far, we have only used the score. Since we have a segmentation and every score note is associated with a performance note\footnote{In reality, not every score note is associated with a performance note since the pianist may have missed some notes. These notes will be ignored} we can also define expression per constituent. Three parameters, analogous to YQX's targets, will be used to describe expression per constituent.

The segmentation, score features and expression parameters are based only on the \textit{melody} of the piece. In this case, melody is defined to be the highest notes in the top staff.

The resulting data is used to train a first order hidden Markov model. The system uses this model to generate performances given a score. To do this, the score is segmented into constituents, score features are extracted for each constituent. Finally viterbi decoding is used to find the sequence of expression parameters that best explains the observed score features.

The simplification of ignoring note level expression is of course unjustified and severly limits the expressiveness of the performances that can be generated. However, if succesful, the resulting performances will clearly demonstrate the phenomenon of structure level expression. A succesful performance rendering system should incorporate both structure and note level expression, section \ref{integration} will explore two possibilities for an integration of SBPR and note level performance rendering.

The succes of a SBPR system depends largely on two factors. The ability to generate musically meaningful parse trees of a piece and the ability to accurately characterize the individual constituents and their relations with other constituents in score features. The following section address these issues.

\begin{itemize}
\item Mention GTTM
\item Give short overview of the system?
\end{itemize}

@@

\section{Method}

This section will describe individual components of the system sketched in section \ref{approach} in more detail. The structural analysis of is based on the delta framework, which will be described in section \ref{deltaframework}.  Section \ref{segmentation} will discuss how the delta framework is applied to get a segmentation. Sections \ref{features} and \ref{targets} will describe how the scorefeatures and expression parameters are calculated. Section \ref{model} will describe how a hidden Markov model is trained on our data.

\subsection{The Delta Framework}

In his Phd thesis \ref{markwin}, Markwin van der Berg introduces a formal way of parsing music into parsetrees: the delta framework. He relates his work to the work of Lerdahl and Jackendoff \ref{lerdahl1996generative} but claims to have found a more general approach. Below, I will shortly describe the deltaframework as proposed by Van der Berg.

@@

\[M_{ij} = [n_i, n_{i+1}, \cdots, n_j]\]

A set of basic features, $\Phi$, is assigned to each note. These are: \textbf{onset}, \textbf{pitch}, \textbf{loudness} and \textbf{release} (called offset by Van den Berg). From these two other features can be derived: \textbf{duration} and \textbf{virtual duration}. Duration is defined as \textbf{release($n_i$)} - \textbf{onset($n_i$)} while virtual duration is defined as \textbf{onset($n_{i+1}$)} - \textbf{onset($n_i$)}.

The basic note features can be used to define delta functions, for example $\Delta\textbf{Pitch} = \textbf{Pitch}(n_i) - \textbf{Pitch}(n_{i-1})$. In general, a delta function $\delta(i)$ is defined as the difference of two notes in some feature $\phi$: $\delta(i) = \phi(n_i) - \phi(n_{i-1})$. We can apply a delta function to every pair of succeeding notes in a sequence to get a list of deltas: 
\[\Delta M_{ij} = [\delta(n_{i+1}), \delta(n_{i+2}), \cdots, \delta(n_{j})]\]

A recursive rule, called the delta rule can be used to translate a list of deltas into a grouping structure. The deltarule, DeltaRule($M_{ij}$), where $M_{ij}$ is the ordered list of notes that is to be analyzed, is defined as follows\footnote{The version here is a slightly reformulated version of Van den Berg's deltarule}:
\begin{algorithm}
\begin{algorithmic}
\STATE{$D_{i+1,j} \leftarrow \Delta M_{ij}$}
\STATE{$A \leftarrow []$}
\STATE $m \leftarrow \textbf{ max}(D)$
\FOR{$\delta \textbf{ in } D$}
\IF{$\delta = m$}
\STATE $p = \textbf{ index of }\delta \textbf{ in } D$
\STATE $q = \textbf{ index of the next occurence of m in }D \textbf{ or } j$
\STATE $\textbf{append DeltaRule(}M_{pq}\textbf{) to } A$
\ENDIF
\ENDFOR
\RETURN{$A$}
\end{algorithmic}
\caption{The delta rule}
\label{deltarule}
\end{algorithm}

It also possible to define higher order delta function (deltas of deltas). Second order deltafunction can be used to differentiate a group of ascending notes from a group of notes that have wildly varying pitches. Van den Berg notes that this generates ambigous grouping structures as for example a second order deltarule needs at least three notes to be specified. For three notes, three different grouping structures are possible: $[[n_1, n_2], n_3]$, $[n_1, n_2, n_3]]$ and $[n_1, [n_2, n_3]]$. Because we want to use a second order deltafunction for segmentation only, we choose to only use one interpretation: $[n_1, n_2], n_3]$. EXPLAIN WHY?
\subsection{Segmentation}
\label{segmentation}

@@

\subsection{Constituent Features}
\label{scorefeatures}

\begin{itemize}
\item We are talking about SCOREFEATURES here, mention this
\end{itemize}

We can now convert a piece of music into a serie of constituents. These constituents will be used to predict expression so we must be able to charaterize them in a way correlates with the way they are performed. Analogous to YQX we are looking for the \textit{context} of the constituent as well as some description of the constituent itself.

YQX uses a set of three score features: pitch interval, duration ratio and I-R arch. The pitch interval is simply the difference in pitch between the current note and the next note. The duration ratio is the logarithmic ratio of the duration of the current note and the duration of the next note.  The I-R arch is is the distance to the nearest point of closure, where closure is calculated from the Implication-Realization analysis. (Too literally?)

We can generalize pitch interval and duration ratio per note to constituent features: \textit{average pitch interval} and \textit{average duration ratio}. Definitions can be found in table \ref{features}. Since I-R arch is related to note-level expression it does not generalize well to a constituent level feature.

The two features above provide information about the constituent context: if they are both zero the constituent is apparently similar in average pitch and average duration. At note level there is not much to say about the current note besides the pitch and duration. However at constituent we would also like to say something about the constituent itself. For this purpose \textit{average delta pitch} and \textit{average delta duration} are used. These features say something about the amount change in pitch and the amount of change in note duration.

The complete set of score features consists of two \textit{context features} and two \textit{constituent features}.

\begin{table}
\begin{itemize}
\item[\textbf{Average pitch interval}] 
The difference between the average pitch of the current constituent and the average pitch of the next constituent, zero if the current is the last constituent
\item[\textbf{Average duration ratio}]
The logarithmic ratio between the average note duration of the current constituent and the average note duration of the next constituent
\end{itemize}
\caption{Context features}
\label{features}
\end{table}

\begin{table}
\begin{itemize}
\item[\textbf{Average delta pitch}] The average of all absolute pitch intervals within one constituent. 
\item[\textbf{Average delta duration}] The average of all absolute differences in duration of succeeding notes within one constituent
\end{itemize}
\caption{Constituent Features}
\label{features}
\end{table}

@@

\subsection{Expression Parameters}
\label{targets}

@@

\subsection{Model}

\section{Implementation}
\section{Performance Rendering}
\begin{itemize}
\item Extract melody
\item Extract scorefeatures from melody
\item lookup notes in deviation, ignore missing notes
\item Extract expressive and non-expressive melody and extract expression features
\item Perform notes other than melody notes the same as the nearest melody note
\end{itemize}
\subsection{Dataset and representation}
\begin{itemize}
\item attack release
\item tempo curves
\end{itemize}
\subsection{Dealing with polyphony}
\subsection{Discretization}

@@

\section{Evaluation and Results}
The proposed system actually requires more training data to do meaningful statistics. However, since expressive interpretation can be 
\subsection{Subjective Listening}
\subsection{Correlation}
\section{Integration with Note Level Performance Rendering}

@@

Unfortunately we do not have acces to the large dataset that YQX uses. The dataset we use is smaller. Since we do not learn per note but per group of notes, the impact of a smaller dataset is even larger. We have discretize into rather large bins for this reason, resulting in cartoonistic performances.

@@

\begin{itemize}
\item Top notes is not always the melody, musical attention
\item Bass and harmony shouldn't be played with the same expression as melodynotes
\item Dataset size
\item Targets for expression level expression
\end{itemize}

\section{Future Work}
\begin{itemize}
\item Combine deltatrees
\item Use loudness and tempo direction instead of averages (requires more data)
\item Generalize approach to incorporate hierarchical structure, let structure extend to note level, so note level expression and structurelevel expression become integrated
\end{itemize}
\subsection{Repetition}
A notion of repetition and similarity would certainly improve the system. Repetition is a very good indicator of constituent breaks. Repetition and similarity could also be used to improve expressiveness of performances. It is probably telling when a phrase is repeated three times and then slightly altered the fourth time. Although finding similarity and musically significant repetition is a subject of its own the delta framework could help to define repetition arbitrarily of transposition or rhythm. A list of pitch deltas can for example be used to detect repetition independend of transposition and a list of duration deltas can be used to detect repetition of rhythm independend of the notes used.
